---
title: "Introduction to the Rout package"
author: "Richard Hooijmaijers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The Rout package is developed to easily create pdf or html reports including tables,
listings and plots created in R. There are a couple of things that might come to mind when you are reading this like `knitr` or `rmarkdown` for instance... These packages do a great job in making very nice reports. I often use them when writing small to medium size documents. I mainly use it to combine written text with at some points some code, plots or tables to help telling the story. There are however times when I am are working in an R script and just want to see how a table or plot look like in a browser or pdf.  
Also in many cases only the results needs to be shared, or someone else will write the report. Furthermore, when writing
large reports with relatively little R results, knitr or rmarkdown might not be the best option.

There are of course options in R to create relatively simple output like the default `png()` and `pdf()` or more advanced like `xtable` or `tables`. Also there are package to directly create output reports like `greport`, `rapport` or `startgazer`. However none of the available options I found satisfied the need that I had for such a system. I hope I can explain in this vignette what my preferred workflow is (as simple as possible) and how I implemented it in the package.

The package has two sets of functions to create documents in either tex or html format. Furthermore
a few functions are present to easily calculate frequencies and descriptive statistics. Finally, the
created results can be displayed in various styles using the template system implemented through
the `whisker` package.

# Calculation functions

To demonstrate the usage of the package the tips dataset from the reshape2 package was used as
this dataset has a nice combination of numerical and discrete values:
```{r}
library(Rout)
library(reshape2)
data(tips)
tips$day <- factor(tips$day,levels=c("Thur","Fri","Sat","Sun" ),ordered = TRUE)
head(tips)
```

Let's start with some simple calculations. Say we want to know basic statistics for the total bill per gender, day of the week and time of day. the following can be done to achieve this:
```{r}
tbl1 <- means(tips,variable="total_bill",by=c("sex","day","time"),total="sex")
head(tbl1,20)
```
The function above calculates a standard set of descriptive statistics frequently used in reporting clinical
results. Also it places the results in a long format data frame (for easy processing in the other functions). There are some options for
calculation of totals and number of digits in output but mainly it uses plyr to do the work.
For this reason a predefined set of statisitcs was chosen as it is quite easy to calculate other statistics just using plyr.

Using a similar function, the frequencies within a data frame can be calculated. Let's say we want
to know the frequencies of each gender, day of the week and time of day within the data frame.
Something like the following can be used:
```{r}
tbl2       <- freq(tips,vars=c("sex","day","time"),total=c("sex","day"),spacechar = "~")
head(tbl2)
```
In general the base of the function is the default `table` within R. Although percentages are
also given. Options are available for the denominator, calculation of totals and taking
duplicates into account. This was mainly inspired by clinical safety analyses.

# Tabular output

To create tabular output using the package, multiple functions can be used. The most simple method is
to use `ltx_doc()` or `html_doc()` to create either a tex (and if compiled a pdf) or html file.
Within these functions a text vector is provided that can include tex/html code and the function will add the
necessary pre and post code to generate a full document.
The second method is by using `ltx_list()` or `html_list()`. These functions take a data frame and places the result in a document. Finally `ltx_table()` and `html_table()` can be used to create various types of (reshaped) tables.

## The doc functions

The doc functions `ltx_doc` and `html_doc` are quite generic in the sense that any kind of
character vector can be included. This means simple text can be added to a document or the
output of functions that can create TeX/html coding. The following example demonstrates how it works

```{r}
ltx_doc("This is just some \\emph{text}")
html_doc("This is just some <em>text</em>")
```

At first sight it might seem that there is little added value in the above function, however
it's also possible to output the results of `xtable` or similar packages. Combine this with an
output statement and you can directly see the compiled pdf (or tex if you like).
The same is demonstrated to create a html file.
The resulting html and pdf files for this and all subsequent examples are available in the links
below the code block. Also in some examples the html version is not displayed but work quite the same as the latex version.

```{r echo=TRUE,eval=FALSE}
library(xtable)
xtbl     <- means(tips,variable="total_bill",by=c("day"))
xtbl_ltx <- print(xtable(xtbl),print.results=FALSE)
ltx_doc(xtbl_ltx,out="out1.tex")

xtbl_html <- print(xtable(xtbl),print.results=FALSE,type="html", html.table.attributes = "class=table")
html_doc(xtbl_html,out="out1.html",
         css="https://maxcdn.bootstrapcdn.com/bootstrap/3.3.6/css/bootstrap.min.css")
```
<a target="_blank" href="out1.pdf">out1.pdf</a><br/>
<a target="_blank" href="out1.html">out1.html</a>


## The list functions

With the list functions you can directly output a data frame without using xtable. The function is intended to display a data frame. Various options can be set within the function to make the output
a bit more attractive. Within the following example the header is be adapted and the variables are grouped and ordered. Furthermore a title, footnote and column width is adapted

```{r echo=TRUE,eval=FALSE}
lst1 <- means(tips,variable="tip",by=c("sex","day"),pack=2)
lst2 <- means(tips,variable="total_bill",by=c("sex","day"),pack=2)
lst  <- merge(lst1,lst2,by=c("sex","day","statistic"))
names(lst)[4:5] <- c("tip","Total Bill")

ltx_list(lst,vargroup=c("","","","income","income"),group = 2, title="a listing",
         fill="-",footnote="a footnote",mancol="p{3cm}llll",out="out2.tex")
```
An html document can be created quite similar, with the exception of the column width option
```{r echo=TRUE,eval=FALSE}
html_list(lst,vargroup=c("","","","income","income"),group = 2,title="a listing",
          fill="-",footnote="a footnote",out="out2.html")
```
<a target="_blank" href="out2.pdf">out2.pdf</a><br/>
<a target="_blank" href="out2.html">out2.html</a>

## The table functions

The most comprehensive functions are `ltx_table` and  `html_table`. With these functions
it is possible to reshape the data before presenting the results. This can make the results
easier to read and offers the possibility to tweak the table header. Because the data is transposed
to a wide format it is important to keep the results "unique". Read more about aggregating in `reshape2::dcast` which is used in this function.The following example demonstrates a common usage of the function:

```{r echo=TRUE,eval=FALSE}
ltx_table(tbl1,x=c("sex","statistic"),y=c("day","time"),var="value",
          title="total bill statistics",xabove=TRUE,out="out3.tex")
html_table(tbl1,x=c("sex","statistic"),y=c("day","time"),var="value",
          title="total bill statistics",xabove=TRUE,out="out3.html")
```
<a target="_blank" href="out3.pdf">out3.pdf</a><br/>
<a target="_blank" href="out3.html">out3.html</a>

The xabove option makes it possible to group the data and to save space. the header has two levels
as two variables are selected to cast the results. Although you can add many levels this way, it is
advised to use no more than three for the sake of readability.
Another examples is presented where labels are added to the data frame and where some common options
are adapted
```{r echo=TRUE,eval=FALSE}
attr(tbl2$sex,"label") <- "Gender"
attr(tbl2$day,"label") <- "Day of the week"
ltx_table(tbl2,x=c("sex","day"),y=c("time"),var="FreqPerc",yhead = TRUE,size="\\normalsize",
          title="customer frequency",xabove=TRUE,out="out4.tex",group=1,mancol="lllll",
          template=paste0(system.file(package =  "Rout"), "/listing.tex"))
```
<a target="_blank" href="out4.pdf">out4.pdf</a>

With a bit of fantasy you can make combinations like this for instance:
```{r echo=TRUE,eval=FALSE}
library(plyr)
tbl3           <- freq(tips,c("sex","day","time"),total=c("day","time"),spacechar = "~")
tbl3$statistic <- factor("N (perc)")
tbl3$value     <- tbl3$FreqPerc
tbl3           <- rbind.fill(tbl1,tbl3)
tmp            <- paste0(system.file(package =  "Rout"), "/listing.tex")
ltx_table(tbl3,x=c("sex","day"),y=c("time","statistic"),var="value",convchar = TRUE,
          title="total bill statistics",xabove=TRUE,out="out5.tex",
          mancol=paste(rep("l",16),collapse=""),
          template=paste0(system.file(package =  "Rout"), "/listing.tex"))
```
<a target="_blank" href="out5.pdf">out5.pdf</a>

# Graphical output

The graphical output system is basically the same as for the tables and listings. The only difference is that the figures are also saved as png or pdf files. Then they are referenced in the tex or html file,
and if selected compiled. Using grid graphics is the easiest because you can just place the plot in an object and pass it to the functions. The way base plots can be used is by placing it in a function.
Some examples are given below:
```{r echo=TRUE,eval=FALSE}
library(ggplot2)
pl <- ggplot(tips,aes(x=day,y=total_bill)) + geom_boxplot()
ltx_plot(pl,out="out6.tex",titlepr = "plot A",title="example for plotting lists")
html_plot(pl,out="out6.html",titlepr = "plot A",title="example for plotting lists")

# you can output a list of plots for instance:
pl <- lapply(unique(tips$sex),function(x){
  ggplot(tips[tips$sex==x,],aes(time,total_bill)) + geom_boxplot() +
    facet_wrap(~day) + ggtitle(x)
})
ltx_plot(pl,out="out7.tex",titlepr = "plot B",title="example for plotting lists")
html_plot(pl,out="out7.html",titlepr = "plot B",title="example for plotting lists")

# base plots can be used but placed in function:
bpl <- function() plot(tips$day,tips$total_bill)
ltx_plot(bpl(),out="out8.tex",titlepr = "plot C",title="example for base plot")
html_plot(bpl(),out="out8.html",titlepr = "plot C",title="example for base plot")
```
<a target="_blank" href="out6.pdf">out6.pdf</a><br/>
<a target="_blank" href="out6.html">out6.html</a><br/>
<a target="_blank" href="out7.pdf">out7.pdf</a><br/>
<a target="_blank" href="out7.html">out7.html</a><br/>
<a target="_blank" href="out8.pdf">out8.pdf</a><br/>
<a target="_blank" href="out8.html">out8.html</a>


# Putting it together

The last thing to after creating all these tables listings and plots is to combine it in an overall
document. Here the last two functions will come in `ltx_combine` and `html_combine`.
To have a bit of background on how these functions work we should first look at all the output that is created. For each table and listing, by default 3 files are generated. For latex output, the tex file,  compiled pdf and a raw tex file. For html only a html and raw html file. The raw files have only the text/coding for the table or listing.
The combine functions can pick these files up and place it in a combined document, and that's basically it.
```{r echo=TRUE,eval=FALSE}
ltx_combine(out="comb1.tex",template=paste0(system.file(package="Rout"),"/default.tex"))
html_combine(out="comb1.html",toctheme=TRUE,
             template=paste0(system.file(package="Rout"),"/bootstrap.html"))
```
<a target="_blank" href="comb1.pdf">comb1.pdf</a><br/>
<a target="_blank" href="comb1.html">comb1.html</a>

# The templates

You probably noticed that not all output generated has the same layout. This is controlled by the templates,
I didn't explain the template argument before because I think it deserves a section of it's own.
The template system uses `whisker` to create output in different styes. A couple
of templates are included within the package to start with, however users can generate their own
templates using this system. The whisker package is set up to provide a template and a list with
variables to render (called rendlist in this package). Provide a template and a rendlist to create
a custom styled output.
The included templates are demonstrated below with the combine functions (but will work the same for the list, table and plot functions). But if you want to experiment read further for some details.

```{r echo=TRUE,eval=FALSE}
tmpl  <- list.files(system.file(package="Rout"),pattern="\\.tex$",full.names=TRUE)
lapply(1:length(tmpl),function(x){
  ppt <- ifelse(grepl("beamer.tex",tmpl[x]),TRUE,FALSE)
  ltx_combine(out=paste0("combt",x,".tex"),template=tmpl[x],presentation=ppt)
})
```
<a target="_blank" href="combt1.pdf">comb1.pdf</a><br/>
<a target="_blank" href="combt2.pdf">comb2.pdf</a><br/>
<a target="_blank" href="combt3.pdf">comb3.pdf</a><br/>
<a target="_blank" href="combt4.pdf">comb4.pdf</a><br/>
<a target="_blank" href="combt5.pdf">comb5.pdf</a><br/>


## Create your own template

The way the package handles the rendlist is that certain information
is always added to this list within a function. This is off course necessary to be able to include the
table,list or plot to be created within the final document. Practically this means that the 'rrres'
tag is always included in the template to place in the R results to be included. Other special tags include
rrtoc, css and orientation. The safest way is to not include these tags in rendlist or change this
within the template (or at least be aware of the effect when omitting tags or placing them elsewhere
in the template)
